---
title: "Lecture 12"
subtitle: "Estimating abundance: Occupancy modeling"
author: "<br/><br/><br/>WILD6900 (Spring 2019)"
output:
  xaringan::moon_reader:
    css: ["default", "WILD6900.css", "WILD6900-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE)
library(WILD6900)
# library(gganimate)
```

## Readings

> Kéry & Schaub 383-409

---
## Estimating abundance? 

#### For the past few weeks, we've been modeling abundance:

$$\Large N_t \sim Poisson(\lambda)$$

--
#### Occupancy is a the proportion of sites that are occupied (alternatively, the probability a site is occupied)

- reduced-information form of abundance
    + If $\large N_i > 0$, $\large z_i = 1$

    + If $\large N_i = 0$, $\large z_i = 0$

--
#### So even when occupancy is the state-variable, we are still modeling something related to abundance

---
## Estimating abundance? 

#### Typical state model for occupancy

$$\Large z_i \sim Bernoulli(\psi)$$

--
#### If the expected abundance is $\large \lambda$, what is the probability $\large N = 0$?

$$\Large Pr(N=0|\lambda)=\frac{\lambda^0e^-\lambda}{0!}=e^{-\lambda}$$ 

--
#### If the expected abundance is $\large \lambda$, what is the probability $\large N > 0$?

$$\Large 1 - Pr(N=0|\lambda)=1 -e^{-\lambda}$$ 

--
#### So (if our assumptions are valid):

$$\Large \psi = 1 - e^{-\lambda}$$

---
## Why estimate occupancy?

#### Less effort

--
#### Historical data sets

--
#### More reliable when $\large N$ very small

--
#### Occupancy = abundance (e.g., breeding territory)

--
#### Metapopulation dynamics

--
#### Distribution/range size

--
#### Disease dynamics

---
## Why not just use observed presence/absence?

#### As in all ecological studies, we rarely (if ever) observe the state process perfectly 

--
#### In the case of occupancy, some sites will be occupied but we will fail to detect the species

- i.e., $\large p < 1$


--
#### Also possible (though generally more rare) that we record the species when it's not present (false positive)

- see [Royle & Link 2006](https://www.uvm.edu/rsenr/vtcfwru/spreadsheets/occupancy/Occupancy%20Exercises/Exercise12/Royle_Link_2006.pdf)


--
#### Similar to N-mixture models, estimating $\Large p$ requires temporal replication

---
## Estimating $\LARGE p$

#### Imagine a single site surveyed 3 times:

- Assume site is closed across samples

- Assume constant $\large p$

$$\Large y_i = [111]$$

--
#### What is the likelihood of this observation?

--
$$\Large \psi p^3$$

--
#### What about?

$$\Large y_i = [011]$$

--
$$\Large \psi (1-p)p^2$$


---
## Estimating $\LARGE p$

#### What about?

$$\Large y_i = [000]$$

--
$$(1 - \psi) + \psi (1-p)^3$$


---
## Single-season (static) occupancy model

#### State-space formulation

- State-model

$$\Large z_i \sim Bernoulli(\psi)$$

$$\Large logit(\psi_i) = \alpha_0 + \alpha_1x_i$$

- Observation-model

$$\Large y_{i,k} \sim Bernoulli(z_ip)$$

$$\Large logit(p_{ik}) = \beta_0 + \beta_1x_{i,k}$$

---
## Single-season (static) occupancy model

```{r eval = FALSE, echo = TRUE}
model{
  # Priors 
  psi ~ dbeta(1, 1)
  p ~ dbeta(1, 1)
  
  # Likelihood
  for(i in 1:M){
    # State model
    z[i] ~ dbern(psi)
    
    # Observation model
    for(k in 1:K){
      y[i, k] ~ dbern(p * z[i])
    }
  }
}
```

---
## Single-season (static) occupancy model

```{r eval = FALSE, echo =TRUE}
model{
  alpha0 ~ dnorm(0, 0.1)
  alpha1 ~ dnorm(0, 0.1)
  mu.p ~ dnorm(0, 0.1)
  tau.p ~ dunif(0, 10)

  for(i in 1:M){
    z[i] ~ dbern(psi[i])
    logit(psi[i]) <- alpha0 + alpha1 * x1[i]
    for(k in 1:K){
      y[i, k] ~ dbern(p[i,k] * z[i])
      logit(p[i,k]) <- lp[i,k]
      lp[i,k] ~ dnorm(mu.lp, tau)
    }
  }

  N.occ <- sum(z[1:M])
}
```


---
## Multi-season (dynamic) occupancy model

#### What if occupancy can change over time? 

- Data collection using the *robust design*
    + Population open between primary periods (e.g., years)
    + Population closed within secondary periods (e.g., occasions)

$$\Large y_i = [\underbrace{101}_{Year\;1}\;\;\;\;\; \underbrace{000}_{Year\;2}\;\;\;\;\;\underbrace{110}_{Year\;3}\;\;\;\;\;\underbrace{100}_{Year\;4}]$$

--
- In year 1:
$$\Large z_{i,1} \sim Bernoulli(\psi)$$

- In years 2+:
$$\Large z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1}\gamma))$$

As always, we can include covariates in $\large \epsilon$ and $\large \gamma$
