---
title: "Quantifying temporary emigration using multi-state models"
author: "WILD6900"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
  - \usepackage{amsmath}
  - \usepackage{blkarray}
  - \usepackage{bigstrut}
  
vignette: >
  %\VignetteIndexEntry{multi-state1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE, warning = FALSE, 
  collapse = TRUE,
  comment = "#>"
)
library(WILD6900)
```

As we discussed in the [open-population capture-recapture lecture lecture](https://rushinglab.github.io/WILD6900/articles/Lecture14/Lecture14.html#1), the CJS model assumes that all emigration is permanent. This assumption is necessary because if we get, for example, a capture history of `101`, we assume that the individual was present and *available to be detected* on the second occasion. We simply failed to detect it and that occurred with probability $1-p_2$. But what if the individual was not available to be detected? Maybe it's breeding territory spanned the study area boundary and it was simply off the plot on the second occasion. Maybe we're studying plants and an individual was dormant for the second year of the study. Or maybe we only sample during the breeding season and this individual skipped breeding in the second year. In all cases, the individual was *temporarily* unavailable for detection. In other words, on the second occasion $p_2 = 0$ for this individual.  

If this temporary emigration is random, estimates of $p$ will be too low (because all the instances where $p=0$ will be incorrectly used to estimate $p$) but estimates of $\phi$ will be unbiased. However, if the temporary emigration process is *Markovian* $-$ that is, whether an individual is unavailable for detection depends on whether the individual was available/unavailable in the previous time step $-$ estimates of survival can be biased. 

Kory & Schaub (2015) provide a nice example of using multi-state capture-recapture models to estimate (and correct for) temporary emigration that is caused by individuals leaving and re-entering the study site. In this model, individuals can be in one of three states: "alive and present", "alive and absent", and "dead". We can define the probabilities associated with each state using three parameters:

- $\phi$: survival probability

- $\psi_{IO}$: probability of moving from **I**n to **O**ut (i.e., from present to absent)

- $\psi_{OI}$: probability of moving from **O**ut to **I**n (i.e., from absent to present)

Based on these three parameters, we can create that state transition matrix:


```{r echo = FALSE}
library(kableExtra)
ms_df <- data.frame(state = c("Present", "Absent", "Dead"),
                 A = c("\\(\\phi (1 - \\psi_{IO})\\)", 
                       "\\(\\phi \\psi_{OI}\\)", 
                       "0"),
                 B = c("\\(\\phi \\psi_{IO}\\)",
                       "\\(\\phi (1 - \\psi_{OI})\\)", 
                       "0"),
                 Dead = c( "\\(1 - \\phi\\)",
                           "\\(1 - \\phi\\)",
                           "1"))

names(ms_df) <- c("", "Present", "Absent", "Dead")
ms_df %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, font_size = 14) 
```

Although individuals can be in three states, there are only two observations $-$ "seen" and "not seen" $-$ and therefore a single detection parameter $p$. Note that $p$ is interpreted as the probability that an individual is captured *given that it is both **alive** and **present** on a given capture occassion*. As a result, we never detect individuals that are "alive and absent" or individuals that are dead:

```{r echo = FALSE}
library(kableExtra)
ms_df2 <- data.frame(state = c("Present", "Absent", "Dead"),
                 seen = c("\\(p\\)", 
                       "0", 
                       "0"),
                 ns = c("\\(1 - p\\)",
                       "1", 
                       "1"))

names(ms_df2) <- c("", "Seen", "Not Seen")
ms_df2 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, font_size = 14) 
```


## Simulated data

Following Kéry & Schaub (2015), we will simulate data as if it came from a capture-recapture study of hibernating salamanders. Individuals are captured and photographed in a single cave where the species hibernates over winter. This cave is not, however, the only hibernation site so when we go out to sample, individuals may be hibernating in a different cave. We will assume the following parameters:


```{r}
# Define mean survival, transitions, recapture, as well as number of occasions, states, observations and released individuals
phi <- 0.85
psiIO <- 0.2
psiOI <- 0.3
p <- 0.7

# Sampling parameters
n.occasions <- 8   # Number of capture occasions
n.states <- 3      # Number of true states
n.obs <- 2         # Number of observed states
```

Next, create matrices to store the transition and observation probabilities:

```{r}
# 1. State process matrix
PSI.STATE <- matrix(c(phi*(1-psiIO), phi*psiIO,     1-phi,
                      phi*psiOI,     phi*(1-psiOI), 1-phi,
                      0,             0,             1), 
                    nrow = n.states, byrow = TRUE)
# 2.Observation process matrix
PSI.OBS <- matrix(c(p, 1-p,
                    0, 1,
                    0, 1), 
      nrow = n.states, byrow = TRUE) 
```

To simulate the multi-state data, we'll use a custom function from the Kéry & Schaub (2015) book. This is a fairly complex function so we won't go through it in detail (though you are encouraged to walk through it on your own. Understanding how the data is simulated will no doubt help you understand the model better). Note that we simulate the `state` and `event` using `rmultinom`. This function takes three arguments: `n`, `size` and `prob`. In this case, we simulate one value (`n=1`) with a single trial (`size = 1`; this is equivalent to using a categorical distribution). For `prob`, we provide a vector with the probability of transitioning to each state, in this case using the rows in the state and observation matrices: 

```{r}
# Define function to simulate multistate capture-recapture data
simul.ms <- function(PSI.STATE, PSI.OBS, marked, n.occasions, unobservable = NA){
   CH <- CH.TRUE <- matrix(NA, ncol = n.occasions, nrow = sum(marked))
   # Define a vector with the occasion of marking
   mark.occ <- matrix(0, ncol = dim(PSI.STATE)[1], nrow = sum(marked))
   g <- colSums(marked)
   for (s in 1:dim(PSI.STATE)[1]){
      if (g[s]==0) next  # To avoid error message if nothing to replace
      mark.occ[(cumsum(g[1:s])-g[s]+1)[s]:cumsum(g[1:s])[s],s] <-
      rep(1:n.occasions, marked[1:n.occasions,s])
      } #s
   for (i in 1:sum(marked)){
      for (s in 1:dim(PSI.STATE)[1]){
         if (mark.occ[i,s]==0) next
         first <- mark.occ[i,s]
         CH[i,first] <- s
         CH.TRUE[i,first] <- s
         } #s
      for (t in (first+1):n.occasions){
         # Multinomial trials for state transitions
         if (first==n.occasions) next
         state <- which(rmultinom(n = 1, size = 1, prob = PSI.STATE[CH.TRUE[i,t-1],])==1)
         CH.TRUE[i,t] <- state
         # Multinomial trials for observation process
         event <- which(rmultinom(n = 1, size = 1, prob = PSI.OBS[CH.TRUE[i,t],])==1)
         CH[i,t] <- event
         } #t
      } #i
   # Replace the NA and the highest state number (dead) in the file by 0
   CH[is.na(CH)] <- 0
   CH[CH==dim(PSI.STATE)[1]] <- 0
   CH[CH==unobservable] <- 0
   id <- numeric(0)
   for (i in 1:dim(CH)[1]){
      z <- min(which(CH[i,]!=0))
      ifelse(z==dim(CH)[2], id <- c(id,i), id <- c(id))
      }
   return(list(CH=CH[-id,], CH.TRUE=CH.TRUE[-id,]))
   # CH: capture histories to be used
   # CH.TRUE: capture histories with perfect observation
}
```

Now we're ready to simulate the data. We first create a matrix with 70 individuals captured on each occasion (these individuals are by definition in state 1 "alive and present" on their first capture occasion). The data simulation function will then create capture histories for each individual. After simulating the capture histories, we create a vector with the first capture occasion for each individual and then recode the observed data: 

```{r}
marked <- matrix(NA, ncol = n.states, nrow = n.occasions)
marked[,1] <- rep(70, n.occasions) # Present
marked[,2] <- rep(0, n.occasions)  # Absent
marked[,3] <- rep(0, n.occasions)  # Dead 

# Execute simulation function
sim <- simul.ms(PSI.STATE, PSI.OBS, marked, n.occasions)
CH <- sim$CH

# Compute vector with occasion of first capture
get.first <- function(x) min(which(x!=0))
f <- apply(CH, 1, get.first)

# Currently, "not seen" is coded as 0 but that is not allowed (we can't index a matrix by row 0!)
# So here we recode the values so that:
# 1 = seen alive, 2 = not seen
rCH <- CH  # Recoded CH
rCH[rCH==0] <- 2

```

Multi-state models can be very computationally expensive. One way to speed them up is the provide the latent matrix `z` as partially-observed data. When the individual is seen, we know $z$. When the individual is not seen, we don't know the state so we code that as `z = NA`. The following function will take the observed capture history and create a matrix `z` with the partially-known states:

```{r}
known.state.ms <- function(ms, notseen){
  # notseen: label for not seen
  state <- ms
  state[state==notseen] <- NA
  for (i in 1:dim(ms)[1]){
    m <- min(which(!is.na(state[i,])))
    state[i,m] <- NA
  }
  return(state)
}
```

We can also define a function to create initial values for `z` and thereby avoid the dreaded `Invalid parent node` error:

```{r}
# Function to create initial values for unknown z
ms.init.z <- function(ch, f){
  for (i in 1:dim(ch)[1]){ch[i,1:f[i]] <- NA}
  states <- max(ch, na.rm = TRUE)
  known.states <- 1:(states-1)
  v <- which(ch==states)
  ch[-v] <- NA
  ch[v] <- sample(known.states, length(v), replace = TRUE)
  return(ch)
}

```

## The model

Here's the JAGS code to fit this model. Work through the code so you know what it's doing:

```{r eval = FALSE}
sink("jags/tempem.jags")
cat("
    model {

    # ---------------------------------
    # Parameters:
    # phi: survival probability
    # psiIO: probability to emigrate
    # psiOI: probability to immigrate
    # p: recapture probability
    # ---------------------------------
    # States (S):
    # 1 alive and present
    # 2 alive and absent
    # 3 dead
    # Observations (O):
    # 1 seen
    # 2 not seen
    # ---------------------------------

    # Priors and constraints
    phi ~ dunif(0, 1)
    psiIO ~ dunif(0, 1)
    psiOI ~ dunif(0, 1)
    p ~ dunif(0, 1)


    # Define state-transition and observation matrices
    # Define probabilities of state S(t+1) given S(t)
    ps[1,1] <- phi * (1-psiIO)
    ps[1,2] <- phi * psiIO
    ps[1,3] <- 1-phi
    ps[2,1] <- phi * psiOI
    ps[2,2] <- phi * (1-psiOI)
    ps[2,3] <- 1-phi
    ps[3,1] <- 0
    ps[3,2] <- 0
    ps[3,3] <- 1

    # Define probabilities of O(t) given S(t)
    po[1,1] <- p
    po[1,2] <- 1-p
    po[2,1] <- 0
    po[2,2] <- 1
    po[3,1] <- 0
    po[3,2] <- 1


    # Likelihood
    for (i in 1:nind){
    # Define latent state at first capture
    z[i,f[i]] <- y[i,f[i]]
    for (t in (f[i]+1):n.occasions){
    # State process: draw S(t) given S(t-1)
    z[i,t] ~ dcat(ps[z[i,t-1],])
    # Observation process: draw O(t) given S(t)
    y[i,t] ~ dcat(po[z[i,t],])
    } #t
    } #i
    }
    ",fill = TRUE)
sink()
```
### Running the model

Now bundle the data and set the MCMC settings. Currently, `ni` and `nb` are set very low. This model takes a long time to run so these settings will allow us to work out any bugs without having to wait a long time to see the error message. Once you have the model running, increase these values to, say, 50,000 and 10,000:

```{r}
# Bundle data
jags.data <- list(y = rCH, f = f, n.occasions = dim(rCH)[2], nind = dim(rCH)[1], z = known.state.ms(rCH, 2))

# Initial values
inits <- function(){list(mean.phi = runif(1, 0, 1), mean.psiIO = runif(1, 0, 1), mean.psiOI = runif(1, 0, 1), mean.p = runif(1, 0, 1), z = ms.init.z(rCH, f))}  

# Parameters monitored
parameters <- c("mean.phi", "mean.psiIO", "mean.psiOI", "mean.p")

# MCMC settings (change these once you know the model is running)
ni <- 500
nt <- 10
nb <- 100
nc <- 3
```

Finally, fit the model:

```{r eval = FALSE}
# Call JAGS from R
tempemi <- jagsUI::jags(data = jags.data, inits = inits, parameters.to.save = parameters, 
                        model.file = here::here("jags/tempem.jags"), 
                        n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb, 
                        parallel = TRUE)

print(tempemi, digits = 3)
```

```{r include = FALSE}
# Call JAGS from R
tempemi <- jagsUI::jags(data = jags.data, inits = inits, parameters.to.save = parameters, 
                        model.file = here::here("inst/jags/tempem.jags"), 
                        n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb, 
                        parallel = TRUE)
```

Be sure to check the model for convergence and to determine whether the parameter estimates are consistent with the data generating values.
