---
title: "Quantifying temporary emigration using multi-state models"
author: "WILD6900"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
  - \usepackage{amsmath}
  - \usepackage{blkarray}
  - \usepackage{bigstrut}
  
vignette: >
  %\VignetteIndexEntry{multi-state1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE, warning = FALSE, 
  collapse = TRUE,
  comment = "#>"
)
library(WILD6900)
```

As we discussed in the [open-population capture-recapture lecture lecture](https://rushinglab.github.io/WILD6900/articles/Lecture14/Lecture14.html#1), the CJS model assumes that all emigration is permanent. This assumption is necessary because our model assumes that if an individual is alive and present at the site, the probability that we detect the individual is $p$. If emigration is not permanent, that is if individuals temporarily leave the site, they are technically alive and "present" at our site but the probability we detect them is 0. 

If this temporary emigration is random, estimates of $p$ will be too low (because all the instances where $p=0$ will be incorrectly 

```{r echo = FALSE}
library(kableExtra)
ms_df <- data.frame(state = c("Present", "Absent", "Dead"),
                 A = c("\\(\\phi (1 - \\psi_{IO})\\)", 
                       "\\(\\phi \\psi_{OI}\\)", 
                       "0"),
                 B = c("\\(\\phi \\psi_{IO}\\)",
                       "\\(\\phi (1 - \\psi_{OI})\\)", 
                       "0"),
                 Dead = c( "\\(1 - \\phi\\)",
                           "\\(1 - \\phi\\)",
                           "1"))

names(ms_df) <- c("", "Present", "Absent", "Dead")
ms_df %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, font_size = 14) 
```

```{r echo = FALSE}
library(kableExtra)
ms_df2 <- data.frame(state = c("Present", "Absent", "Dead"),
                 seen = c("\\(p\\)", 
                       "0", 
                       "0"),
                 ns = c("\\(1 - p\\)",
                       "1", 
                       "1"))

names(ms_df2) <- c("", "Seen", "Not Seen")
ms_df2 %>%
  kable("html", align = 'c', escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, font_size = 14) 
```


```{r}
# Define function to simulate multistate capture-recapture data
simul.ms <- function(PSI.STATE, PSI.OBS, marked, unobservable = NA){
   # Unobservable: number of state that is unobservable
   n.occasions <- dim(PSI.STATE)[4] + 1
   CH <- CH.TRUE <- matrix(NA, ncol = n.occasions, nrow = sum(marked))
   # Define a vector with the occasion of marking
   mark.occ <- matrix(0, ncol = dim(PSI.STATE)[1], nrow = sum(marked))
   g <- colSums(marked)
   for (s in 1:dim(PSI.STATE)[1]){
      if (g[s]==0) next  # To avoid error message if nothing to replace
      mark.occ[(cumsum(g[1:s])-g[s]+1)[s]:cumsum(g[1:s])[s],s] <-
      rep(1:n.occasions, marked[1:n.occasions,s])
      } #s
   for (i in 1:sum(marked)){
      for (s in 1:dim(PSI.STATE)[1]){
         if (mark.occ[i,s]==0) next
         first <- mark.occ[i,s]
         CH[i,first] <- s
         CH.TRUE[i,first] <- s
         } #s
      for (t in (first+1):n.occasions){
         # Multinomial trials for state transitions
         if (first==n.occasions) next
         state <- which(rmultinom(1, 1, PSI.STATE[CH.TRUE[i,t-1],,i,t-1])==1)
         CH.TRUE[i,t] <- state
         # Multinomial trials for observation process
         event <- which(rmultinom(1, 1, PSI.OBS[CH.TRUE[i,t],,i,t-1])==1)
         CH[i,t] <- event
         } #t
      } #i
   # Replace the NA and the highest state number (dead) in the file by 0
   CH[is.na(CH)] <- 0
   CH[CH==dim(PSI.STATE)[1]] <- 0
   CH[CH==unobservable] <- 0
   id <- numeric(0)
   for (i in 1:dim(CH)[1]){
      z <- min(which(CH[i,]!=0))
      ifelse(z==dim(CH)[2], id <- c(id,i), id <- c(id))
      }
   return(list(CH=CH[-id,], CH.TRUE=CH.TRUE[-id,]))
   # CH: capture histories to be used
   # CH.TRUE: capture histories with perfect observation
}
```

```{r}
# Define mean survival, transitions, recapture, as well as number of occasions, states, observations and released individuals
phi <- 0.85
psiIO <- 0.2
psiOI <- 0.3
p <- 0.7
n.occasions <- 8  
n.states <- 3
n.obs <- 2
marked <- matrix(NA, ncol = n.states, nrow = n.occasions)
marked[,1] <- rep(70, n.occasions)	# Present
marked[,2] <- rep(0, n.occasions)	# Absent
marked[,3] <- rep(0, n.occasions)	# Dead 

# Define matrices with survival, transition and recapture probabilities
# These are 4-dimensional matrices, with 
   # Dimension 1: state of departure
   # Dimension 2: state of arrival
   # Dimension 3: individual
   # Dimension 4: time
# 1. State process matrix
totrel <- sum(marked)
PSI.STATE <- array(NA, dim=c(n.states, n.states, totrel, n.occasions-1))
for (i in 1:totrel){
   for (t in 1:(n.occasions-1)){
      PSI.STATE[,,i,t] <- matrix(c(
      phi*(1-psiIO), phi*psiIO,     1-phi,
      phi*psiOI,     phi*(1-psiOI), 1-phi,
      0,             0,             1       ), nrow = n.states, byrow = TRUE)
      } #t
   } #i

# 2.Observation process matrix
PSI.OBS <- array(NA, dim=c(n.states, n.obs, totrel, n.occasions-1))
for (i in 1:totrel){
   for (t in 1:(n.occasions-1)){
      PSI.OBS[,,i,t] <- matrix(c(
      p, 1-p,
      0, 1,
      0, 1    ), nrow = n.states, byrow = TRUE)
      } #t
   } #i

# Execute simulation function
sim <- simul.ms(PSI.STATE, PSI.OBS, marked)
CH <- sim$CH

# Compute vector with occasion of first capture
get.first <- function(x) min(which(x!=0))
f <- apply(CH, 1, get.first)

# Recode CH matrix: note, a 0 is not allowed!
# 1 = seen alive, 2 = not seen
rCH <- CH  # Recoded CH
rCH[rCH==0] <- 2

```

```{r}
known.state.ms <- function(ms, notseen){
  # notseen: label for ënot seení
  state <- ms
  state[state==notseen] <- NA
  for (i in 1:dim(ms)[1]){
    m <- min(which(!is.na(state[i,])))
    state[i,m] <- NA
  }
  return(state)
}

# Function to create initial values for unknown z
ms.init.z <- function(ch, f){
  for (i in 1:dim(ch)[1]){ch[i,1:f[i]] <- NA}
  states <- max(ch, na.rm = TRUE)
  known.states <- 1:(states-1)
  v <- which(ch==states)
  ch[-v] <- NA
  ch[v] <- sample(known.states, length(v), replace = TRUE)
  return(ch)
}

```

```{r}
# Bundle data
jags.data <- list(y = rCH, f = f, n.occasions = dim(rCH)[2], nind = dim(rCH)[1], z = known.state.ms(rCH, 2))

# Initial values
inits <- function(){list(mean.phi = runif(1, 0, 1), mean.psiIO = runif(1, 0, 1), mean.psiOI = runif(1, 0, 1), mean.p = runif(1, 0, 1), z = ms.init.z(rCH, f))}  

# Parameters monitored
parameters <- c("mean.phi", "mean.psiIO", "mean.psiOI", "mean.p")

# MCMC settings
ni <- 500
nt <- 10
nb <- 100
nc <- 3

# Call WinBUGS from R (BRT 35 min)
tempemi <- jagsUI::jags(data = jags.data, inits = inits, parameters.to.save = parameters, 
                        model.file = here::here("inst/jags/tempem.jags"), 
                        n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb, 
                        parallel = TRUE)

print(tempemi, digits = 3)
```