---
title: "Introduction to random effects and hierarchical models"
author: "WILD6900"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
  - \usepackage{amsmath}
  
vignette: >
  %\VignetteIndexEntry{poisson_glm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
library(WILD6900)
```

In this activity, we will use the process of simulating data to understand what random effects are and how they are interpreted in hierarchical models.

***
**Objectives**

* Simulate data sets with both fixed and random effect structure   

* Fit fixed and random effects models in JAGS  

***

For this lab, let's assume we're interested in modeling the the body mass (g) of lizards as a function of body length (mm). We capture lizards on $J$ study plots and measure the mass and length of each individual. 

## The model

In this example, we have a single continuous predictor variable and $J$ levels of a factor. We'll assume for now that the residual body mass of each individual is normally distributed after accounting for body length. This suggests that following model:

$$y_{ij} = \alpha_{[j]} + \beta \times x_i + epsilon_i$$

$$\alpha_{[j]} \sim normal(\mu_{alpha}, \tau_{alpha})$$

$$\epsilon_i \sim normal(0, \tau)$$

First, set up the simulation parameters:

```{r eval = FALSE}
J <- 20               # Number of sites
N <- 200              # Number of observations
Length <- rnorm(N)    # Scaled body length

## Randomly determine each individuals study plot
plot <- sample(x = 1:J, size = N, replace = TRUE)
table(plot)
```

```{r echo = FALSE}
J <- 20               # Number of sites
N <- 200              # Number of observations
Length <- rnorm(N)    # Scaled body length

## Randomly determine each individuals study plot
plot <- sample(x = 1:J, size = N, replace = TRUE)
knitr::kable(table(plot))
```

Now, let's set the parameters in the model: 

```{r}
mu.alpha <- 23   # Overall mean body mass
sigma.alpha <- 2 # standard deviation of group-level means
tau.alpha <- 1/sigma.alpha^2

alpha <- rnorm(J, mu.alpha, sigma.alpha)

beta <- 6     # Slope of body length on mass
sigma <- 3    # Residual standard deviation
tau <- 1/sigma^2
```

To create the linear predictor, we'll use the `model.matrix()` function to create the model matrix corresponding to our model and then multiply it by our parameters:

```{r}
Xmat <- model.matrix(~as.factor(plot) + Length - 1)
Xmat[1:5,c(1,2,3,21)]
```

Now use matrix multiplicaiton to create the linear predictor:

```{r}
linear.pred <- Xmat %*% c(alpha, beta)
head(linear.pred)
```

And finally generate the observations and combine the relevant vectors within a single data frame:

```{r fig.width=6, fig.height=6}
mass <- rnorm(N, mean = linear.pred, sd = sigma)
mass_df <- data.frame(plot = as.factor(plot), length = Length, pred_length = linear.pred, mass = mass)

ggplot(mass_df, aes(x = length, y = mass, color = plot)) +
  geom_abline(intercept = alpha, slope = beta) +
  geom_point() +
  scale_color_brewer() +
  guides(color = "none")
```


## Fitting the model

Now we're ready to fit the model in JAGS. Code for this model can be accessed with:

```{r}
model.file <- system.file("jags/random_ancova.jags", package = "WILD6900")
```

Next, prepare the data, initial values, and MCMC settings. Notice the need to generate $J$ starting values of $\alpha$:

```{r eval = FALSE}
jags_data <- list(y = mass_df$mass, N = nrow(mass_df), 
                  J = J,
                  plot = plot,
                  x = mass_df$length)

jags_inits <- function(){list(mu.alpha = rnorm(1),
                              sigma.alpha = runif(1),
                              alpha = rnorm(J), # Notice we need J initial values 
                              beta = rnorm(1),
                              tau = runif(1))}

params <- c("mu.alpha", "sigma.alpha", "alpha", "beta", "tau", "sigma")

re_fit <- jagsUI::jags(data = jags_data, inits = jags_inits,
                    parameters.to.save = params, 
                    model.file = model.file,
                    n.chains = 3, n.iter = 10000, n.burnin = 2500, n.thin = 1)

print(re_fit)
```

```{r include = FALSE}
jags_data <- list(y = mass_df$mass, N = nrow(mass_df), 
                  J = J,
                  plot = plot,
                  x = mass_df$length)

jags_inits <- function(){list(mu.alpha = rnorm(1),
                              sigma.alpha = runif(1),
                              alpha = rnorm(J), # Notice we need J initial values 
                              beta = rnorm(1),
                              tau = runif(1))}

params <- c("mu.alpha", "sigma.alpha", "alpha", "beta", "tau", "sigma")

re_fit <- jagsUI::jags(data = jags_data, inits = jags_inits,
                    parameters.to.save = params, 
                    model.file = model.file,
                    n.chains = 3, n.iter = 10000, n.burnin = 2500, n.thin = 1)
```

```{r echo = FALSE}
knitr::kable(re_fit$summary[1:10,], format = 'html')
```
Just for kicks, let's fit the same data to a model that assumes fixed-effects on the $\alpha$'s:

```{r eval = FALSE}
model.file <- system.file("jags/fixed_ancova.jags", package = "WILD6900")

jags_inits <- function(){list(alpha = rnorm(J), # Notice we need J initial values 
                              beta = rnorm(1),
                              tau = runif(1))}

params <- c("alpha", "beta", "tau", "sigma")

fe_fit <- jagsUI::jags(data = jags_data, inits = jags_inits,
                    parameters.to.save = params, 
                    model.file = model.file,
                    n.chains = 3, n.iter = 10000, n.burnin = 2500, n.thin = 1)

print(fe_fit)
```

```{r include = FALSE}
model.file <- system.file("jags/fixed_ancova.jags", package = "WILD6900")

jags_inits <- function(){list(alpha = rnorm(J), # Notice we need J initial values 
                              beta = rnorm(1),
                              tau = runif(1))}

params <- c("alpha", "beta", "tau", "sigma")

fe_fit <- jagsUI::jags(data = jags_data, inits = jags_inits,
                    parameters.to.save = params, 
                    model.file = model.file,
                    n.chains = 3, n.iter = 10000, n.burnin = 2500, n.thin = 1)

print(fe_fit)
```

```{r echo = FALSE}
knitr::kable(fe_fit$summary[1:10,], format = 'html')
```

Finally, combine results from the two models to visualize differences in the $\alpha$ estimates:

```{r fig.width=6, fig.height=6}

compare_mods <- data.frame(alpha_mean = c(alpha, re_fit$mean$alpha, fe_fit$mean$alpha),
                           CI_2.5 = c(rep(NA, J), re_fit$q2.5$alpha, fe_fit$q2.5$alpha),
                           CI_97.5 = c(rep(NA, J), re_fit$q97.5$alpha, fe_fit$q97.5$alpha), 
                           n = c(table(plot), table(plot) - 0.25, table(plot) + 0.25),
                           model = rep(c("True", "Random", "Fixed"), each = J))

ggplot(compare_mods, aes(x = alpha_mean, color = model, y = n)) + 
  geom_vline(xintercept = mu.alpha, color = "grey50", linetype = "longdash") +
  geom_vline(xintercept = re_fit$mean$mu.alpha, color = "#CD0200", alpha = 0.75) +
  geom_errorbarh(aes(xmin = CI_2.5, xmax = CI_97.5), height = 0) +
  geom_point(size = 4, color = "white") +
  geom_point() +
  scale_color_manual(values = c("#446E9B", "#D47500", "black")) +
  scale_x_continuous(expression(alpha)) +
  scale_y_continuous("Sample size")
```