---
title: "Lecture 8"
subtitle: "Introduction to hierarchical models"
author: "<br/><br/><br/>WILD6900 (Spring 2021)"
output:
  xaringan::moon_reader:
    css: ["default", "WILD6900.css", "WILD6900-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
library(WILD6900)
library(gganimate)
```

## Readings

> Kéry & Schaub 73-82

---
class: inverse, center, middle

# What are random effects?

---
## What are random effects?
<br/>

--
- Fixed effects are constant across observational units, random effects vary across units  
<br/>

--
- Fixed effects are used when you are interested in the specific factor levels, random effects are used when you are interested in the underlying population  
<br/>

--
- When factors levels are exhaustive, they are fixed. When they are a sample of the possible levels, they are random  
<br/>

--
- Random effects are the realized values of a random variable  
<br/>

--
- Fixed effects are estimated by maximum likelihood, random effects are estimated with shrinkage  

???

Definitions from Gelman & Hill (2007) pg. 245

---
## A simple linear model

$$\Large y_{ij} = \beta_{[j]} + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$

--
- If $\beta_{[1]} = \beta_{[2]} = \beta_{[3]} = ...=\beta_{[J]}$

--
```
model {
  # Priors
  beta0 ~ dnorm(0, 0.33)
  tau ~ dgamma(0.25, 0.25)
    
  # Likelihood
  for (i in 1:N){
    y[i] ~ dnorm(mu[i], tau)        
    mu[i] <- beta0
  } #i
}
```

---
## A simple linear model

$$\Large y_{ij} = \beta_{[j]} + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$

--
- If $\beta_{[1]} \perp \beta_{[2]} \perp \beta_{[3]} \perp ...\perp \beta_{[J]}$

--
```
model {
  # Priors
  for(j in 1:J){
    beta0[j] ~ dnorm(0, 0.33)
  }
  tau ~ dgamma(0.25, 0.25)
    
  # Likelihood
  for (i in 1:N){
    y[i] ~ dnorm(mu[i], tau)        
    mu[i] <- beta0[group[j]]
  } #i
}
```
???

nb $\perp$ means "independent of"

This should look familiar - it's the means parameterization of the ANOVA model

---
## A simple linear model

$$\Large y_{ij} = \beta_{[j]} + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$

--
- If $\beta_{[j]} \sim normal(\mu_{\beta0}, \tau_{\beta0})$

```{r, fig.height = 5, fig.width = 8}
x <- seq(from = 0, to = 20, length.out = 200)

df <- data.frame(x = x, density = dnorm(x, mean = 10, sd = 2))

beta0 <- c(4, 8, 10, 11, 13)


ggplot(df, aes(x = x, y = density))  +
  scale_x_continuous("", breaks = beta0, 
                     label = c(expression(beta[0[1]]), 
                               expression(beta[0[2]]),
                               expression(beta[0[3]]),
                               expression(beta[0[4]]),
                               expression(beta[0[5]]))) +
  geom_segment(aes(x = beta0[1], xend = beta0[1], y = -Inf, yend = dnorm(beta0[1], 10, 2)), color = "grey60") +
  geom_segment(aes(x = beta0[2], xend = beta0[2], y = -Inf, yend = dnorm(beta0[2], 10, 2)), color = "grey60") +
  geom_segment(aes(x = beta0[3], xend = beta0[3], y = -Inf, yend = dnorm(beta0[3], 10, 2)), color = "grey60") +
  geom_segment(aes(x = beta0[4], xend = beta0[4], y = -Inf, yend = dnorm(beta0[4], 10, 2)), color = "grey60") +
  geom_segment(aes(x = beta0[5], xend = beta0[5], y = -Inf, yend = dnorm(beta0[5], 10, 2)), color = "grey60") +
  geom_path(size = 2, color = WILD6900_colors$value[WILD6900_colors$name=="primary"])
```

---
## A simple linear model

$$\Large y_{ij} = \beta_{[j]} + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$

--
- If $\beta_{[j]} \sim normal(\mu_{\beta0}, \tau_{\beta0})$

--
```
model {
  # Priors
  for(j in 1:J){
    beta0[j] ~ dnorm(mu.beta, tau.beta)
  }
  mu.beta ~ dnorm(0, 0.33)
  tau.beta ~ dgamma(0.25, 0.25)
  tau ~ dgamma(0.25, 0.25)
    
  # Likelihood
  for (i in 1:N){
    y[i] ~ dnorm(mu[i], tau)        
    mu[i] <- beta0[group[j]]
  } #i
}
```

---
## Random effects
<br/>

--
- Only apply to factors  
<br/>

--
- Imply grouped effects  
<br/>

--
- Can include intercept, slope, and variance parameters  
<br/>

--
- Assume exchangeability  
<br/>

--
- Represent a compromise between total pooling $(\beta_{0[1]}=\beta_{0[2]}=...=\beta_{0[J]})$ and no pooling $(\beta_{[1]} \perp \beta_{[2]} \perp ...\perp \beta_{[J]})$  
<br/>

--
- Typically require $>5-10$ factor levels  


---
## Random effects = hierarchical model
<br/>
<br/>


$$\Large [\beta_{0[j]}, \mu_{\beta0}, \tau_{\beta0}, \tau|y_{ij}] = [y_{ij}|\beta_{0[j]}, \tau][\beta_{0[j]}|\mu_{\beta0}, \tau_{\beta0}][\tau][\mu_{\beta0}][\tau_{\beta0}]$$
<br/>

--
- Can include multiple random effects  

- Can include fixed and random effects (mixed-models)  

- Can include multiple levels of hierarchy  

---
## Why use random effects?

1) Scope of inference  

- learn about $\beta_{0[j]}$ **and** $\mu_{beta_0}$, $\tau_{beta_0}$  

- prediction to unsampled groups (in space or time)  


---
## Why use random effects?

1) Scope of inference  

2) Partitioning of variance  

- Account for variability among groups vs. among observational units

---
## Why use random effects?

1) Scope of inference  

2) Partitioning of variance  

3) Accounting for uncertainty  

- modeling $\tau_{beta_0}$ recognizes uncertainty from sampling groups

---
## Why use random effects?

1) Scope of inference  

2) Partitioning of variance  

3) Accounting for uncertainty  

4) Avoiding psuedo-replication  

- Account for non-independence within groups

---
## Why use random effects?

1) Scope of inference  

2) Partitioning of variance  

3) Accounting for uncertainty  

4) Avoiding psuedo-replication  

5) Borrowing strength  

- $\beta_{0[j]}$ estimating from group $j$ observations + all other groups  

- "shrinkage"" towards mean  

    + degree of shrinkage inversely proportional to precision  
    
---
## Why not use random effects?

#### Always use random effects (Gelman & Hill 2007)

#### but...

--
- Assumption of exchangeability  
<br/>

--
- Requires 5-10 levels  
<br/>

--
- Computationally intensive  
<br/>

--
- Harder to interpret  
