---
title: "Lecture 8"
subtitle: "Introduction to hierarchical models"
author: "<br/><br/><br/>WILD6900 (Spring 2019)"
output:
  xaringan::moon_reader:
    css: ["default", "WILD6900.css", "WILD6900-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE)
library(WILD6900)
library(gganimate)
```

## Readings

> Kéry & Schaub 73-82

```{r}
N <- 200               # Number of observations
J <- 5                 # Number of study plots

mu.beta0 <- 20
sigma.beta0 <- 4
beta0 <- rnorm(J, mu.beta0, sigma.beta0)

mu.beta1 <- 0.8           # Relationship between body length and body mass
sigma.beta1 <- 0.2
beta1 <- rnorm(J, mu.beta1, sigma.beta1)

sigma2 <- 1.2            # Process variance
tau <- 1/sqrt(sigma2)  # Precision 

length <- rnorm(N)     # Body lengths

x <- rmultinom(N, size = 1, prob = c(0.03, 0.3, 0.3, 0.2, 0.17))
plot <- apply(x, 2, function(x) which(x == 1))

mass <- data.frame(ID = seq(1:N),
                   plot = as.factor(plot),
                   length = length)

mass <- dplyr::mutate(mass, pred.mass = beta0[plot] + beta1 * length,
                      mass = rnorm(N, pred.mass, sqrt(sigma2)))
```

---
class: inverse, center, middle

# What are random effects?

---
## What are random effects?
<br/>

--
- Fixed effects are constant across observational units, random effects vary across units  
<br/>

--
- Fixed effects are used when you are interested in the specific factor levels, random effects are used when you are interested in the underlying population  
<br/>

--
- When factors levels are exhaustive, they are fixed. When they are a sample of the possible levels, they are random  
<br/>

--
- Random effects are the realized values of a random variable  
<br/>

--
- Fixed effects are estimated by maximum likelihood, random effects are estimated with shrinkage  

???

Definitions from Gelman & Hill (2007) pg. 245


---
class: inverse, center, middle

# The question

### How does body length influence body mass?

---
# The data

- $\large y_i$ = body mass of individual $\large i$

- $\large x_i$ = body length of individual $\large i$

- $\large j$ = study plot of individual $\large i$ $\large (j \in [1,2,3,...,J])$

---
# Model 1

<br/>
$$\Large y_i = \beta_0 + \beta_1 \times x_i + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$
<br/>

--
```{r fig.width=4.5, fig.height=4.5}
ggplot(mass, aes(x = length, y = mass)) + geom_point() +
  geom_abline(slope = mean(beta1), intercept = mean(beta0), size = 2, color = WILD6900_colors$value[WILD6900_colors$name=="warning"])
```

---
## Assumptions of model 1

--
- Basic assumptions of linear regression
<br/>

--
- The mean body mass 

---
# Model 2

---
# Model 1

<br/>
$$\Large y_{ij} = \beta_{0[j]} + \beta_1 \times x_i + \epsilon_i$$

$$\Large \epsilon_i \sim normal(0, \tau)$$
<br/>

--
```{r fig.width=4.5, fig.height=4.5}
ggplot(mass, aes(x = length, y = mass)) + geom_point(aes(color = plot)) +
  geom_abline(slope = mean(beta1), intercept = beta0, size = 2)
```

